---
title: "Programming with R"
subtitle: "EPSY 887 - Fall 2019"
author: "Jason Bryer, Ph.D."
institute: "University at Albany"
date: 2019-09-19
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, my-theme.css]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r echo=FALSE, message=FALSE, warning=FALSE, messages=FALSE, results='hide'}
library(ggplot2)
library(psych)
library(reshape2)
library(readxl)
library(tidyverse)
library(DATA606)

mass1 <- read_xlsx('../datasets/MathAnxiety.xlsx') # Summer 2014
mass2 <- read_xlsx('../datasets/MathAnxiety.xlsx', sheet = 2) # Fall 2014
mass3 <- read_xlsx('../datasets/MathAnxiety.xlsx', sheet = 3) # Summer 2015
mass4 <- read_xlsx('../datasets/MathAnxiety.xlsx', sheet = 4) # Fall 2019
items <- c('I find math interesting.',
		   'I get uptight during math tests.',
		   'I think that I will use math in the future.',
		   'Mind goes blank and I am unable to think clearly when doing my math test.',
		   'Math relates to my life.',
		   'I worry about my ability to solve math problems.',
		   'I get a sinking feeling when I try to do math problems.',
		   'I find math challenging.',
		   'Mathematics makes me feel nervous.',
		   'I would like to take more math classes.',
		   'Mathematics makes me feel uneasy.',
		   'Math is one of my favorite subjects.',
		   'I enjoy learning with mathematics.',
		   'Mathematics makes me feel confused.')

mass1$Term <- 'Summer 2014'
mass2$Term <- 'Fall 2014'
mass3$Term <- 'Summer 2015'
mass4$Term <- 'Fall 2019'
mass <- rbind(mass1, mass2, mass3, mass4)

set.seed(2112)
```


# if Statements

`if` statements are a way of doing different operations conditionally. In this example, the `sample` function returns either a 0 or 1.

```{r}
if(sample(c(0,1), 1) > 0.5) {
	print('heads')
} else {
	print('tails')
}
```

For simpler problems, the `ifelse` function may be easier.

```{r}
ifelse(sample(c(0,1), 1) > 0.5, 'heads', 'tails')
```

---

# for loops

The `for` loop allows you to do some operation a number of times. You define some variable `i` that is set to somve value, then the all the R code within the braces (i.e. `{ }`) will be executed.

```{r}
for(i in 2:5) {
	print(i)
}
```

---

# Using for loop to recode

In this example, we wish to convert each of the Likert responses in the `mass` data frame to an ordered factor.

```{r}
for(i in 2:15) {
	mass[,i] <- factor(mass[,i], levels=1:5, labels=c('Strongly Disagree', 
			'Disagree', 'Neutral', 'Agree', 'Strongly Agree'), ordered=TRUE)
}
str(mass)
```

---

# The apply functions

There are a number of `Xapply` functions that behave like `for` loops. They perform an operation over some data structure.

* `apply` - Used to execute a function to each row (`MARGIN = 1`) or column (`MARGIN = 2`).
* `lapply` - Used to execute a function to each element in a list. Returns a list of the same length of what was passed in.
* `sapply` - Used to execute a function to each element in a list. Returns a vector or array of the same length of what was passed in. This only works if the function executed on each element returns a vector of of the same length for all results.
* `vapply` - Is similar to sapply, but has a pre-specified type of return value, so it can be safer (and sometimes faster) to use.
* `tapply` - Apply a function to each cell of a ragged array, that is to each (non-empty) group of values given by a unique combination of the levels of certain factors.
* `mappy` - Is a multivariate version of sapply. mapply applies FUN to the first elements of each ... argument, the second elements, the third elements, and so on. 

---

# Examples

```{r}
apply(mass1[,2:15], 2, mean, na.rm = TRUE)
```

--

```{r}
x <- list(a = 1:10, beta = exp(-3:3), logic = c(TRUE,FALSE,FALSE,TRUE))
```

.pull-left[

```{r}
lapply(x, mean)
```

]
.pull-right[

```{r}
sapply(x, mean)
sapply(x, quantile)
```
]

---


# while loops

`while` loops will run until some condition is met. In this example, consider 1=heads, 0=tails. How many random events would it take before getting 100 heads?

```{r}
sum <- 0
count <- 0
while(sum < 100) {
	sum <- sum + sample(c(0,1), 1)
	count <- count + 1
}
count # Number of loops until we got 100 head
```

---

# Functions

Functions allow for organizing common procedures to easily be used later. You can specify any number of parameters with optional default values. The objects created and/or edited within the function are local to the function (i.e . not available to the calling environment). The `return` or `invisible` function is used to return value(s) to the environment calling the function. The `invisible` function will return a value but will not print it to the console if it is not assigned to an object.

```{r}
getStanding <- function(credits, 
						breaks=c(30, 60, 90),
						labels=c('Freshman','Sophomore', 'Junior', 'Senior')) {
	results <- cut(credits, breaks=c(-Inf, breaks, Inf), labels=labels, 
				   include.lowest=TRUE, right=FALSE)
	return(results)
}
getStanding(42)
getStanding(c(27, 60, 101, 200))
```

---

# The ... Operator

The `...` operator in R (technically an ellipsis, but often referred to as the dots parameter, or dot-dot-dot) allows for flexibility in how parameters are given to R. There are two major ways the `...` parameter is used:

1. To pass parameters to another function without having to explicitly define them.
2. To have an arbitrary number of parameters for a function.

---

# ... : Passing Parameters


```{r, eval=TRUE}
print_percent <- function(x, multiply = all(x <= 1.0), ...) {
	if(multiply) {
		x <- x * 100
	}
	x <- round(x, ...)
	return( paste0(x, '%') )
}
```

--

```{r}
print_percent(.2112)
```

--

```{r}
print_percent(.2112, digits = 1)
```

--

```{r}
print_percent(c(.4, .01, .05, .2112), digits = 2)
```

---

# ... : Arbitrary Arugements


```{r}
test_function <- function(..., 
		  param1, 
		  param2 = 'Some Default') {
	dots <- list(...)
	print('The ... parameters:')
	print(dots)
	if(missing(param1)) {
		print('param1 was not defined')
	} else {
		print(paste0('param1 = ', param1))
	}
	print(paste0('param2 = ', param2))
}
```


---

# ... : Arbitrary Arugements (cont)

```{r}
test_function()
```

--

```{r}
test_function(param1 = 2112)
```

---

# ... : Arbitrary Arugements (cont)

```{r}
test_function(a = 'a', b = 'b', 'c', param1 = 2112)
```

---

# Object Oriented Programming in R

Object-oriented programming (OOP) is a way of programming where you define not only the data structure (attributes), but also the operations (actions). Consider a *football*. This is an object that *attributes* (e.g. size, color, shape) and *actions* (e.g. throw, catch, kick).

R has a number of OOP frameworks, however we will only consider the S3 OOP framwork since it is part of base R and you have already encountered it many times. It is an ad hoc OOP framwork in that actions are determined informally. **Generic** functions will dispatch to the class specific function using the class name. Consider the implementation of the `print` function:

```{r}
print
```

The `UseMethod` will determine the correct `print` method to call for the specified object. Consider an object where `class(object) <- c('MyClass')`. Calling `print(object)` will first look for a function `print.MyClass`, if that is not found then it will use the `print.default` function.

---

# Inheritance

A key feature of OOP frameworks is inheritance. Using our ball analogy, we could have a *super* class of type *ball*. It can define some features and actions. But we could also create a class *football* that inherits from *ball*, whereby we "inherit" certain attributes and actions, but can override and/or define new ones. 

Consider the shape attribute of ball. We could represent that a simply the diamater since this is common for most balls. However, footballs are oblong and therefore require two diamaters.

A baseball could also inherit from ball, but we would override the *kick* action since that is not defined for baseball. 

---

# OOP Example

```{r, eval=TRUE}
baseball <- function(...) {
	ball <- list(...)
	class(ball) <- c('baseball', 'ball')
	return(ball)
}

football <- function(...) {
	ball <- list(...)
	class(ball) <- c('football', 'ball')
	return(ball)
}

kick <- function(x, ...) {
	UseMethod("kick")
}

kick.default <- function(x, ...) {
	print("Kicking the ball like any other ball!")
}
```

---

# OOP Example (cont).

```{r}
my_baseball <- baseball()
my_football <- football()
kick(my_baseball)
kick(my_football)
```

---

# OOP Example (cont).

```{r}
kick.baseball <- function(x, ...) {
	print("You don't kick baseballs!")
}

kick(my_baseball)
kick(my_football)
```

---

# OOP Example (cont).

```{r}
kick.football <- function(x, ...) {
	print("Kicking for field goal, or 1-point conversion?")
}

kick(my_baseball)
kick(my_football)
```

---

# The trouble with tibbles

Consider the `mass` data frame, it has the following class definition: `r class(mass)`. Recall that the `[` operator on data frames is a function that has one important parameter, `drop`. 

.pull-left[

```{r}
class(mass)
is.data.frame(mass)
```

]
.pull-right[

We will explicitly convert `mass` to a data.frame.
```{r}
mass_df <- as.data.frame(mass)
class(mass_df)
is.data.frame(mass_df)
```
]
---

# The trouble with tibbles (cont.)

Consider the situation where we wish to retrieve one column:

.pull-left[

```{r}
mass[,1]
class(mass[,1])
```
]

--

.pull-right[

```{r}
mass_df[,1]
class(mass_df[,1])
```

]

---

# Extended Example: Merging Multiple data.frames

See: https://github.com/jbryer/EPSY887-Fall2019/blob/master/R/merge_all.R

---

# Bootstrapping

* First introduced by Efron (1979) in [*Bootstrap Methods: Another Look at the Jackknife*](https://projecteuclid.org/euclid.aos/1176344552).
* Estimates confidence of statistics by resampling *with* replacement.
* The *bootstrap sample* provides an estimate of the sampling distribution.
* The `boot` R package provides a framework for doing bootstrapping: https://www.statmethods.net/advstats/bootstrapping.html

---

# Bootstrapping Example (Population)

Define our population with a uniform distribution.

```{r}
n <- 1e5
pop <- runif(n, 0, 1)
mean(pop)
```

```{r, echo=FALSE, fig.width=12,fig.height=5}
d <- density(pop)
h <- hist(pop, plot=FALSE)
hist(pop, main='Population Distribution', xlab="", freq=FALSE, 
     ylim=c(0, max(d$y, h$density)+.5), col=COL[1,2], border = "white", 
	 cex.main = 1.5, cex.axis = 1.5, cex.lab = 1.5)
lines(d, lwd=3)
```

---

# Bootstrapping Example (Sample)

We observe one random sample from the population.

```{r}
samp1 <- sample(pop, size = 50)
```

```{r, echo=FALSE, fig.width=12,fig.height=5}
d <- density(samp1)
h <- hist(samp1, plot=FALSE)
hist(samp1, main='Distribution of Sample', xlab="", freq=FALSE, 
     ylim=c(0, max(d$y, h$density)+.5), col=COL[1,2], border = "white", 
	 cex.main = 1.5, cex.axis = 1.5, cex.lab = 1.5)
lines(d, lwd=3)
```

---

# Bootstrapping Example (Estimate)

```{r}
boot.samples <- numeric(1000) # 1,000 bootstrap samples
for(i in seq_along(boot.samples)) { 
	tmp <- sample(samp1, size = length(samp1), replace = TRUE)
	boot.samples[i] <- mean(tmp)
}
head(boot.samples)
```

---

# Bootstrapping Example (Distribution)

```{r, fig.width=12,fig.height=5}
d <- density(boot.samples)
h <- hist(boot.samples, plot=FALSE)
hist(boot.samples, main='Bootstrap Distribution', xlab="", freq=FALSE, 
     ylim=c(0, max(d$y, h$density)+.5), col=COL[1,2], border = "white", 
	 cex.main = 1.5, cex.axis = 1.5, cex.lab = 1.5)
lines(d, lwd=3)
```

---

# 95% confidence interval

```{r}
c(mean(boot.samples) - 1.96 * sd(boot.samples), 
  mean(boot.samples) + 1.96 * sd(boot.samples))
```

---

# Bootstrapping is not just for means!

```{r}
boot.samples.median <- numeric(1000) # 1,000 bootstrap samples
for(i in seq_along(boot.samples.median)) { 
	tmp <- sample(samp1, size = length(samp1), replace = TRUE)
	# NOTICE WE ARE NOW USING THE median FUNCTION!
	boot.samples.median[i] <- median(tmp) 
}
head(boot.samples.median)
```

95% confidence interval for the median

```{r}
c(mean(boot.samples.median) - 1.96 * sd(boot.samples.median), 
  mean(boot.samples.median) + 1.96 * sd(boot.samples.median))
```

---

# Your Turn: Create a Boostrap Function

The following R code should run successfully:

```{r, eval = FALSE}
# Setup our data
n <- 1e5
pop <- runif(n, 0, 1)
samp1 <- sample(pop, size = 50)

# Create an object that did the bootstrap samples	
boot1 <- bootstrap(samp1)
class(boot1) # Verify the class is "bootstrap"
# Call some S3 generic functions
print(boot1)
print(boot1, ci = 0.95)
summary(boot1)
summary(boot1, median)
summary(boot1, sd)
plot(boot1)
plot(boot1, metric = median)
```

---

# Define the required functions

Write the function that will create our object, and define our class (use `bootstrap` as the class name). The function will need (at least) two parameters: `x` which will be the sample from which the bootstrap distribution will be estimated, and `n.boot.samples` which defines the number of bootstrap samples to generate. You can optionally include the `...` parameter, but we won't use it here.

--

```{r, eval = FALSE}
bootstrap <- function(x, n.boot.samples = 1000, ...) {
	boot <- list()
	# Need to so some work here...
	class(boot) <- 'bootstrap'
	return(boot)
}
```

--

Define S3 functions/methods for `print`, `summary`, and `plot`:

--

```{r, eval = FALSE}
print.bootstrap <- function(x, ...) { 
}

summary.bootstrap <- function(x, ...) {
}

plot.bootstrap <- function(x, ...) {
}
```

---

# Bootstrap Function

Implement the `bootstrap` function such that it creates a `list` with two elements:
* `samples` - a list of length `n.boot.samples` where each element is a bootstrap sample (i.e. a vector with length equal to `length(x)`).
* `sample` - the value of `x`.

--

```{r}
bootstrap <- function(x, 
					  n.boot.samples = 1000,
					  ...) {
	bootstrap <- list()
	bootstrap$samples <- list()
	for(i in seq_len(n.boot.samples)) { 
		bootstrap$samples[[i]] <- sample(x, size = length(x), replace = TRUE)
	}
	bootstrap$sample <- x
	class(bootstrap) <- 'bootstrap'
	return(bootstrap)
}
```

---

# Summary Function

Implement the `summary` function so that it returns the bootstrap value. The user should be able to define what metric to use (e.g. `mean`, `median`, `sd`, etc.)

--

```{r}
summary.bootstrap <- function(x, metric = mean, ...) {
	boot_dist <- sapply(x$samples, metric, ...)
	return(mean(boot_dist))
}
```

---

# Print Function

Implement the `print` function so that it prints the following information to the console: (1) the number of bootstrap samples, (2) the bootstrap value, and (3) a confidence interval.

--

```{r}
print.bootstrap <- function(x, metric = mean, ci, ...) {
	boot_dist <- sapply(x$samples, metric, ...)
	value <- mean(boot_dist)
	value.sd <- sd(boot_dist)
	ci.str <- ''
	if(!missing(ci)) {
		cv <- abs(qnorm((1 - ci) / 2))
		ci.str <- paste0(' (', round(ci * 100), '% confidence interval: ',
						 round(value - cv * value.sd, digits = 2), ', ', 
						 round(value + cv * value.sd, digits = 2), ')')
	}
	print(paste0('Bootstrap value of ', round(value, digits = 2), ci.str,
			 ' was obtained from a bootstrap distribution (n = ', length(x$samples), 
			 ') of an original sample of ', length(x$sample), ' observations.'))
}
```

---

# Plot Function

Implement the `plot` function so that it plots a histogram of the bootstrap distributions.

--

```{r}
plot.bootstrap <- function(x, metric = mean, ...) {
	boot_dist <- sapply(x$samples, metric, ...)
	d <- density(boot_dist)
	h <- hist(boot_dist, plot=FALSE)
	hist(boot_dist, main='Bootstrap Distribution', xlab="", freq=FALSE, 
		 ylim=c(0, max(d$y, h$density)+.5), col='#569BBDC0', border = "white", 
		 cex.main = 1.5, cex.axis = 1.5, cex.lab = 1.5)
	lines(d, lwd=3)
}
```

---

# Putting it all together

```{r}
boot1 <- bootstrap(samp1)
class(boot1) # Verify the class is "bootstrap"
print(boot1)
print(boot1, ci = 0.95)
summary(boot1)
summary(boot1, median)
```

---

```{r, fig.width = 10, fig.heigth = 3}
plot(boot1)
```

---

```{r, fig.width = 10, fig.heigth = 3}
plot(boot1, metric = median)
```

